Forest canopies are better generators of biodiversity in the tropics

### 2025-06-27 manuscript.Rmd
ms revisions

For centuries, ecologists have been intrigued by the role of vertical landscape structures in biogeographic patterns from local to global scales (cite). In 1799, Alexander von Humbolt began an expedition that would ultimately canonize the role of mountains in global richness patterns [@vonhumboltEssayGeographyPlants2013; @HumboldtsLegacy2019]. As vertical features on the landscape, mountains host an extraordinarly high biodiversity by increasing the land surface within a given area and producing steep climate gradients that promote high turnover in species composition relative to adjacent climatically homogeneous lowlands [@rahbekHumboldtsEnigmaWhat2019]. Yet, mountains are better generators of biodiversity in the tropics because high latitude mountains have lower elevation snow lines, which truncate the area of suitable habitat, and more variable climates that increase range size and reduce elevational turnover in composition relative to climatically stable tropical mountain slopes (cite Rahbek, Janzen, other mountain paper, and Humbolt essay on geo of plants). Over a century later, MacArthur highlighted the role of another vertical dimension in promoting high biodiversity: the vertical structure of vegetation. Yet only recently has the three-dimensional nature of terrestrial landscapes been recognized as critical to our understanding of biogeographic patterns as our ability to measure this structure and quantify its use by animals globally has increased due to the collation of large ecological databases and launch of satellite missions to measure vegetation structure.

@@

# Tall canopies generate biodiversity, but are better generators of biodiversity in the tropics

@@

>>>>

For centuries, ecologists have been intrigued by the role of vertical landscape features in biogeographic patterns from local to global scales [@vonhumboltEssayGeographyPlants2013; @macarthurBirdSpeciesDiversity1961; @bashamVerticalStratificationPatterns2023a; @xingEcologicalPatternsProcesses2023a]. In 1799, Alexander von Humbolt began an expedition that would ultimately canonize the role of mountains in global richness patterns [@vonhumboltEssayGeographyPlants2013; @HumboldtsLegacy2019]. As vertical, topographically complex landscape features, mountains provide a larger volume of geographic space and produce steep climate gradients over short distance that allow rapid rates of turnover and an extraordinarily high biodiversity for a given area relative to flat and climatically homogeneous lowlands [@rahbekHumboldtsEnigmaWhat2019]. Yet, mountains are better generators of biodiversity in the tropics because high latitude mountains have lower elevation snow lines, which truncate the area of suitable habitat, and fewer climate zones than tropical mountains [@rahbekHumboldtsEnigmaWhat2019; @vonhumboltEssayGeographyPlants2013]. Over a century later, Robert MacArthur highlighted the role of another vertical dimension in promoting high biodiversity: the vertical structure of vegetation [@macarthurBirdSpeciesDiversity1961]. Yet only recently has the three-dimensional nature of terrestrial landscapes been recognized as a critical dimension to our understanding of biogeographic patterns [@fredstonReimaginingSpeciesMove2025; @xingEcologicalPatternsProcesses2023a] as our ability to measure vegetation structure and quantify its use by animals globally has increased due to the collation of large ecological databases and launch of satellite missions to measure forest height and structural complexity [@popkinSpaceLaserMap2018; @mouraPhylogenyinformedCharacterisationGlobal2024a].

@@

# Tall canopies generate biodiversity, but are better generators of biodiversity in the tropics
- general richness patterns
  - increase with area, productivity, and topographic complexity
- mountain biodiversity:
  - tropical regions contain more species than expected for a given area, while temperate and boreal regions contain fewer species than expected for a given area
- our results
  - high latitude forests have lower richness than expected for a given forest height based on a global model of richness by canopy height

@@

- comparison to mountains
  - climate bands are narrower in the tropics -> higher turnover rates
  - we don't see this pattern for vertical climate profiles (similar shapes in profiles across latitudes and less variation in the troipcs)
- results
  - the vertical difference in rhmin and tmax is greater than the latitudinal difference from tropical to boreal forests
  - stable microclimates in tropical canopies -> specialization and speciation
